---
title: "Illustrating and interpreting a FAIR assessment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Illustrating and interpreting a FAIR assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, dpi = 96, fig.align = "center")
library(rfuji)
```

This vignette shows how to *read* the output of an assessment: the scorecard
plot, the score tables, the maturity levels, and the reuse/access context that
`rfuji` adds on top of the F-UJI metrics. For how the scores are computed see
`vignette("methodology")`; for a quick tour see `vignette("rfuji")`.

So the vignette renders offline and deterministically, it uses the bundled
example assessment `fair_example` (a real assessment of a Zenodo deposit,
\doi{10.5281/zenodo.8347772}). You produce your own the same way:

```{r eval = FALSE}
# (needs network) assess any DOI / PID / URL:
x <- assess_fair("https://doi.org/10.5281/zenodo.8347772")
```

```{r}
data(fair_example)
x <- fair_example
x
```

The printed summary is the fastest read: an earned/total and percentage per
FAIR category, the overall score, and any reuse/access/identifier flags.

## 1. The scorecard plot

`plot()` turns the assessment into a one-glance scorecard. Each bar is a FAIR
category filled to its score, labeled with `earned/total` and its **maturity**
level (the colored word on the right). The dark bar at the top is the overall
FAIR score.

```{r fig.alt = "FAIR scorecard: horizontal bars for Findable, Accessible, Interoperable and Reusable, plus an overall FAIR bar."}
plot(x)
```

To see *which* of the 17 metrics drive each category, plot the per-metric
breakdown. Bars are grouped and colored by category (F/A/I/R) and labeled with
the metric identifier and its `earned/total`.

```{r fig.height = 5.6, fig.alt = "Per-metric FAIR breakdown: one horizontal bar per metric, grouped by category."}
plot(x, type = "metric")
```

## 2. Score tables

`summary()` returns the per-category table behind the scorecard (handy for
reports and further computation):

```{r}
summary(x)
```

`as.data.frame()` gives one row per metric, with its principle, category,
score, maturity, and pass/fail status:

```{r}
df <- as.data.frame(x)
head(df, 8)
```

Because it is a plain data frame you can slice it however you like, for example
the metrics that did not earn full marks:

```{r}
df[df$earned < df$total, c("metric_identifier", "metric_name", "earned", "total")]
```

## 3. How to read the numbers

* **Score (`earned`/`total`).** Each metric is worth a fixed number of points;
  the category score is the sum of earned over total across its metrics, and the
  overall FAIR score is the sum across all 17 metrics.
* **Percent.** `earned / total * 100`, shown on each bar.
* **Maturity (FAIR level).** A CMMI level from 0 to 3 (`incomplete`, `initial`,
  `moderate`, `advanced`) summarizing *how far up* the testing ladder a metric
  reached. A metric can earn points yet still sit at a low maturity if only its
  easiest test passed. Maturity is the colored tag on the category scorecard.
* **A low score is a finding, not a verdict.** A restricted-access or
  unlicensed object can be perfectly legitimate; the score tells you what a
  machine could and could not verify from the metadata.

## 4. The context rfuji adds beyond the score

A single FAIR percentage hides *why* an object is or is not reusable. `rfuji`
surfaces that separately (see `vignette("beyond-fuji")`); the same information
is in the assessment object and worth showing alongside the scorecard.

License **reusability** (not merely presence): a license can be detected yet not
actually permit reuse.

```{r}
x$reuse$licenses[[1]][c("license", "category", "rdp_category")]
```

Access level and sensitivity flags (a restricted object is not a FAIR failure,
but you should know):

```{r}
x$access[c("access", "controlled_access", "sensitive")]
```

Identifier hygiene (does the persistent identifier resolve cleanly, no obvious
problems):

```{r}
x$identifier_hygiene[c("scheme", "is_persistent", "hygiene_ok")]
```

## 5. Exporting the illustration

The assessment serializes for downstream tools. `as_fuji_json()` emits a payload
matching the upstream F-UJI `FAIRResults` schema:

```{r}
js <- as_fuji_json(x)
substr(js, 1, 220)
```

`as_rdf()` emits a machine-readable rating (W3C DQV plus a schema.org `Rating`
as JSON-LD), suitable for embedding in a landing page:

```{r}
rdf <- as_rdf(x)
substr(rdf, 1, 220)
```

## Summary

* `plot(x)` and `plot(x, type = "metric")` are the quickest way to see an
  assessment.
* `summary(x)` and `as.data.frame(x)` give the numbers as tidy tables.
* Read score, percent, and maturity together; treat low scores as questions,
  not verdicts.
* The reuse, access, and identifier-hygiene elements explain the *why* behind
  the number.
