---
title: "Getting started with rfuji"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rfuji}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`rfuji` assesses how well a research data object satisfies the FAIR principles
(Findable, Accessible, Interoperable, Reusable), entirely in R. It is a native
port of the [F-UJI](https://github.com/pangaea-data-publisher/fuji) metrics, so
it needs no external assessment server.

```{r}
library(rfuji)
```

## Assessing an object

Pass any DOI, persistent identifier, or URL to `assess_fair()`. It resolves the
identifier, harvests metadata, and scores it against the FAIRsFAIR metrics.

```{r}
a <- assess_fair("https://doi.org/10.5281/zenodo.8347772")
a
```

The returned `fair_assessment` object prints an F/A/I/R summary. The numbers come
from up to 17 metrics; each is one row of:

```{r}
as.data.frame(a)
```

`summary(a)` gives the per-principle score table, and the *maturity* column
reports a 0–3 CMMI level (incomplete → advanced).

```{r}
summary(a)
```

## Interpreting beyond the score

Automated FAIR scores have well-known blind spots. `rfuji` surfaces three:

```{r}
# A license being *present* does not mean the data is open for reuse.
a$reuse                    # per-license: open / restrictive, commercial, derivatives

# Restricted access can be legitimate (e.g. sensitive human data) and should not
# be read as "not FAIR".
a$access                   # access level, controlled_access, sensitive

# Identifiers should follow best practices.
a$identifier_hygiene       # layered / non-persistent identifier warnings
```

You can call these directly too:

```{r}
license_reuse("https://creativecommons.org/licenses/by-nc-nd/4.0/")
identifier_hygiene("RRID:MGI:5577054")
fair_principles()          # canonical FAIR principle definitions
```

## Exporting results

```{r}
as_fuji_json(a)            # F-UJI-compatible FAIRResults JSON
as_rdf(a)                  # W3C DQV + schema.org Rating (JSON-LD)
```

## Interactive use

```{r}
launch_rfuji()             # Shiny app
```

A no-install browser version is at <https://choxos.github.io/rfuji/app/>; because
browsers cannot fetch landing pages cross-origin, it scores from registry
metadata (DataCite/Crossref) only, so some metrics are lower than the R engine.
